**The Heart Lies: The Role of Interoception and Theory of Mind in Deception**

### 2022-11-10 paper/manuscript.Rmd
manu

In this post-truth era, deception has become increasingly prevalent in our everyday lives [@lewandowsky2017beyond]. Although the act of deceiving is regarded as complex and often requires effortful self-regulation to conceal [@abe2007deceiving; @krokoszinski2016emotion], research shows that people tend to under-estimate their ability to successfully lie [@elaad2003effects]. However, relative to the abundant literature regarding deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], surprisingly little attention has been conferred to investigating the inter-individual differences in people's ability to lie. Given meta-analytic findings wherein the deceptive ability of the liar is suggested to be more important than the lie detector in deceptive interactions [@verigin2019lie; @bond2008individual; @levine2011sender], it is perhaps more pertinent to investigate what makes someone a better liar.

@@

Alternatively, interoceptive abilities could also influence inter-individual differences in lying. Broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], @garfinkel2015knowing's conceptualizes interoception as a three-dimensional framework comprising three distinct facets, namely - *interoceptive accuracy*, the objective ability to monitor internal bodily signals; *interoceptive sensibility*, the subjective confidence in one's interoceptive accuracy; and *interoceptive awareness*, the metacognitive awareness of one's interoceptive accuracy (i.e., accuracy-confidence correspondence). Interoception has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Although little research interest has been conferred to investigating the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. In fact, this is consistent with the somatic marker hypothesis, which posits interoception is associated with a propensity towards making rational decisions [@damasio1996somatic].

In contrast, some studies have instead found heightened interoceptive attention, one's self-focus towards internal bodily signals, to predict apathetic, immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. Extending these findings to social cognition, @vabba2022interoceptive further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high (relative to low), whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told.

While the literature pertaining to interoception seems contradictory, it is possible that these mixed findings are a result of a failure to take into consideration interoception's multidimensionality. Indeed, previous interoception research has largely been focused on examining interoceptive accuracy [@wallman2021physical]. Taking this to perspective with the paucity in studies linking deceptive ability with interoception, more studies are warranted to comprehensively examine the relationship between interoception and deception.

@@

To assess participants' interoceptive ability, participants completed a *Heartbeat Counting Task (HCT)* [@schandry1981heart] while having their actual heartbeats recorded. During the HCT task, participants were instructed to count the number of heartbeats they felt over 5 trials with varying time intervals (20s, 25s, 30s, 35s, 40s), the sequence of which was randomized. Interoceptive accuracy (*HCT Accuracy*) was then computed based on the absolute difference between the actual number of heartbeats and the number of heartbeats counted during each trial. To assess their interoceptive sensibility (*HCT Confidence*), participants were asked to rate, on a visual analog, how confident they were of their heartbeat counting accuracy at the end of each interval.

@@

In the first phase of the study, participants answered a brief demographic survey before being administered a battery of questionnaires (*BES*, *MAIA-2*, *Lie Scale*) and cognitive-behavioural tasks (*HCT*, *Yoni Task*). This series of tasks and questionnaires were administered in a randomized order [**TBC**] in order to control for order effects. In the second phase of the study, the *Deception Task* was conducted using PsychoPy [**REF**]. Each participant was conferred both conditions of the *Deception Task*, the order sequence of which was randomized. Each task condition comprised 40 trials, of which participants were instructed to lie in half of the trials (20 **Truth** trials; 20 **Lie** trials). In total, each participant underwent 80 *Deception Task* trials.

@@

While there is an extensive body of literature associating deception with social cognition and emotion regulation, most research had been focused on their roles in deception detection, leaving a paucity in studies pertaining to their influence on lying ability. The present study hence seeks to address this research gap, further aiming to examine the interplay effect of TOM and interoception on our ability to lie convincingly. Notably, we found a significant difference in participants' lying ability (as measured by response time, heart rate and confidence ratings) between the *Polygraph* and *Interrogation* conditions. This hence confers support for the manipulation used in our experimental paradigm, demonstrating that increasing the attentional focus towards external (visual) interoceptive cues can enhance interoceptive sensitivity, therefore improving the precision of internal signals perceived. Extending the growing body of literature pertaining to the application of interoceptive training to improve emotion regulation [@sugawara2020effects; @quadt2021interoceptive; @wu2022enhanced], our study thus provides evidence that interoceptive sensitivity can be manipulated and thereby modified to guide our emotional responses.

@@

<!-- TOM -->
Our results show a significant association between individuals' composite TOM score with lower lie confidence and longer response times when interoceptive cues were highly salient. Recent studies linking interoceptive abilities and social cognition with one's experience of self posit that an accurate sense of internal bodily signals is integral for individuals to maintain a distinction between self and other, while correctly inferring the mental states of others [@shah2017heart; @palmer2018going]. Taking this to perspective with the predictive coding framework [@friston2010free], when interoceptive cues are highly salient, the resulting increase in interoceptive awareness will be conflicted against higher (more precise) exteroceptive TOM. To attenuate this conflict so as to maintain balance between our interoceptive and exteroceptive states, it is hence possible that individuals experienced a re-direction of their attentional resources from external to internally-focused signals to update higher order beliefs. Consistent with the attentional shift hypothesis [@arnold2019interoception; @hanania2010selective], this re-direction of cognitive resources away from that needed to convincingly tell a lie could thus have led to lower lie confidence and a longer response time.

<!-- Potential Implications  -->

<!-- Limitations and Future Research -->
- NOTE: no significant correlation between HCT accuracy and MAIA dimensions (only HCT confidence sig correlates with MAIA dimensions of awarenss and regulation)

While this research has demonstrated that interoceptive abilities may be used to regulate and conceal deception, the individual's confidence in their lying abilities, is subjective, and may not offer a full-picture of their objective success. Future research should continue to unpack this relationship, examining how confidence and interoceptive regulation is reflected in the actual success of a lie, as well as how individuals confidence in their interoceptive abilities may correspond to decision bias [@garfinkel2015knowing], and propensity to engage in deception [@sip2012if]. Furthermore, future research should expand recent findings on how individual traits correlate with true interoceptive abilities [@pearson2022two], to understand how they may factor into deceptive behaviour.

>>>>

In this post-truth era, deception has become increasingly prevalent in our everyday lives [@lewandowsky2017beyond]. Although the act of deceiving is regarded as complex and often requires effortful self-regulation to conceal [@abe2007deceiving; @krokoszinski2016emotion], research shows that people tend to under-estimate their ability to successfully lie [@elaad2003effects]. However, relative to the abundant literature regarding deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], surprisingly little attention has been conferred to investigating the inter-individual differences in people's lying ability. Given meta-analytic findings wherein the deceptive ability of the liar is suggested to be more important than the lie detector in deceptive interactions [@verigin2019lie; @bond2008individual; @levine2011sender], it is perhaps more pertinent to investigate what makes someone a more convincing liar.

@@

Alternatively, interoceptive abilities could also influence inter-individual differences in lying. Broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], @garfinkel2015knowing's conceptualizes interoception as a three-dimensional framework comprising three distinct facets, namely - *interoceptive accuracy*, the objective ability to monitor internal bodily signals; *interoceptive sensibility*, the subjective confidence in one's interoceptive accuracy; and *interoceptive awareness*, one's ability to identify and appropriately respond to their perceived interoceptive states. Interoception has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Although little research interest has been conferred to investigating the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. In fact, this is consistent with the somatic marker hypothesis, which posits interoception is associated with a propensity towards making rational decisions [@damasio1996somatic].

In contrast, some studies have instead found heightened interoceptive attention, one's self-focus towards internal bodily signals, to predict apathetic, immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. Extending these findings to social cognition, @vabba2022interoceptive further reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high (relative to low), whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told. Taking these mixed findings to perspective with the paucity in research linking deceptive ability with interoception, more studies are warranted to comprehensively examine the relationship between interoception and deception.

@@

To assess participants' interoceptive ability, participants completed a *Heartbeat Counting Task (HCT)* [@schandry1981heart] while having their actual heartbeats recorded. During the HCT task, participants were instructed to count the number of heartbeats they felt over 5 trials with varying time intervals (20s, 25s, 30s, 35s, 40s), the sequence of which was randomized. Objective interoceptive accuracy (*HCT Accuracy*) was then computed based on the following formula ($\HCT_{Acc} = 1-|actual-count|/(actual + count)/2)$) [**lol pls help me write this in proper syntax**). To assess their interoceptive sensibility (*HCT Confidence*), participants were asked to rate, on a visual analog scale, how confident they were of their heartbeat counting accuracy at the end of each interval. Interoceptive awareness (*HCT Awareness*) was indexed by the correspondence between one's objective interoceptive accuracy and their confidence ratings.

@@

In the first phase of the study, participants answered a brief demographic survey before being administered a battery of questionnaires (*BES*, *MAIA-2*, *Lie Scale*) and cognitive-behavioural tasks (*HCT*, *Yoni Task*). This series of tasks and questionnaires were administered in a randomized order [**TBC**] in order to control for order effects. In the second phase of the study, the *Deception Task* was conducted using PsychoPy [@peirce2019psychopy2]. Each participant was administered both conditions of the *Deception Task*, the order sequence of which was randomized. Each task condition comprised 40 trials, of which participants were instructed to lie in half of the trials (20 **Truth** trials; 20 **Lie** trials). In total, each participant underwent 80 *Deception Task* trials.

@@

While there is an extensive body of literature associating deception with social cognition and emotion regulation, most research had been focused on their roles in deception detection, leaving a paucity in studies pertaining to their influence on lying ability. The present study hence seeks to address this research gap, further aiming to examine the interplay effect of TOM and interoception on our ability to lie convincingly. Notably, we found a significant difference in participants' lying ability (as measured by response time, heart rate and confidence ratings) between the *Polygraph* and *Interrogation* conditions. This hence confers support for the manipulation used in our experimental paradigm, demonstrating that increasing the attentional focus towards visual interoceptive cues, even if they are fake, can elicit interoceptive awareness and enhance the precision of internal signals if such external cues of interoceptive signals are perceived to be real. Extending previous research pertaining to the application of interoceptive training for improving emotion regulation [@sugawara2020effects; @quadt2021interoceptive; @wu2022enhanced], our study thus provides evidence that interoceptive awareness can be manipulated and thereby modified to guide our emotional responses.

@@

<!-- TOM -->
Our results show a significant association between individuals' composite TOM score with lower lie confidence and longer response time when interoceptive cues were highly salient. Under the free energy principle [@friston2010free], interoception is regarded as integral in TOM [@shah2017heart; @palmer2018going; @ondobaka2017role]. Specifically, during social inference, interoceptive signals are integrated with exteroceptive (e.g., visual and auditory) and proprioceptive information (i.e., motor) to guide one's social behaviour [@ondobaka2017role]. Given recent work in which TOM is associated with one having more precise interoceptive states [@tsakiris2017multisensory; @palmer2018going; @arnold2019interoception], it is possible that individuals with greater TOM abilities have higher interoceptive awareness, more stable self-other representations and less malleable percepts of self [@tsakiris2017multisensory; @arnold2019interoception]. As such, external cues of their interoceptive signals may be perceived as incongruent, hence generating a sense of uncertainty in their responses and leading to their lower lie confidence and longer response times.

<!-- Limitations and Future Research -->
Despite longstanding research interest in deception, little attention has been conferred to understanding the inter-individual differences in lying ability. In this study, we found a significant effect of interoception, particularly pertaining to its metacognitive dimensions, on one's ability to lie convincingly. However, future research is needed to address a few limitations. Namely, objective interoceptive accuracy was only assessed using the HCT. While the HCT is one of the most commonly used measure of objective interoceptive accuracy [@desmedt2022does], concerns regarding its validity has been highlighted in several studies [@desmedt2018heartbeat; @desmedt2022does; @brener2016towards; @legrand2022heart]. Given increasing research efforts invested in the development of novel interoception tasks [@plans2021measuring; @legrand2022heart; @ponzo2021measuring], future works should further examine the relationship between interoception and lying ability using measures with better psychometric properties. Additionally, in order to account for the small sample size, we used factor analysis to reduce the number of dimensions assessing our variables of interest, which could have over-simplified the effect of interoception observed. Taking into perspective that the various facets of interoception are widely regarded as dissociable [@suksasilp2022towards; @mehling2012multidimensional], future studies could thus investigate their effects separately, to better delineate how interoceptive processes influence lying ability.
